Real-time resolution of short-read assembly graph using ONT long reads

### 2020-08-11 plos_latex_edit.tex
more editting

%% END MACROS SECTION

%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}

@@

The first step is to cluster the big anchors (longer than $10$Kbp and in/out degree less than 2) based on their \emph{kmer} coverage.
To achieve this, we applied DBSCAN clustering algorithm \cite{Ester96adensity-based}, which utilises a customised metric function to map contigs into a one-dimensional space.  In order to 
 define a customised metric which is sample and fast to calculate, we assumed that a single long contig itself consists of a Poisson distribution of \emph{k-mers} count with the mean approximated by the contig's coverage. 
%The rationale is to approximate a coverage value of an extremely long contig (which can be split into more than 10,000 \emph{k-mers}) with a sampled mean of a Poisson distribution (of \emph{k-mers} count). 
The metric is then determined by a distance function of two Poisson distributions based on Kullback-Leibler divergence (or relative entropy) between the Poisson distribution representing each contig\cite{Kullback1951information}.

@@

Due to false alignments from shorter contigs to the long reads, not all of the reported step nodes are necessary to be appeared in the ultimate path resolved by the bridge. 
In most cases, the accumulated score of each step indicates its likelihood to be the true component of the final solution.
For that reason, a strategy similar to binary searching is employed to find a path across 2 anchors of a bridge as shown in Algorithm~\ref{algo:binsearch}.

Before that, we define Algorithm~\ref{algo:findpath} to demonstrates the path finding algorithm for two nodes given their estimated distance. In which, function 
$\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n$ 
from line 3 of the algorithm's pseudo code builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance regarding nanopore read error rate) is reached. This task is implemented based on Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

@@

\begin{algorithm}[!hpt]
\DontPrintSemicolon
\KwData{Assembly graph $G\{V,E\}$}
\KwIn{Brigde $B:\lbrace\overrightarrow{v_0}, ...\overrightarrow{v_k}, ...,\overrightarrow{v_n}\rbrace$ with $\overrightarrow{v_0}$ and $\overrightarrow{v_n}$ are two anchors, $\{\overrightarrow{v_k}\}, k=1\ldots(n-1)$ are steps in-between}
\KwOut{Set of candidate paths connecting $\overrightarrow{v_0}$ to $\overrightarrow{v_2}$ that maximize the likelihood of the step list.}
\SetKwFunction{BinaryBridging}{BinaryBridging} 
\SetKwProg{Fn}{Function}{:}{}
\Fn{\BinaryBridging{B}}{
\tcc{search for the contig with maximum score from the step list (two ends excluded)}
$m$:=$\argmax_k(\overrightarrow{v_k}.score())$ \;
\tcc{if there is no step in-between, run path finding algorithm above directly and return the result}
\If{$M.size() \equiv 2$}{ 
	\Return{$\mathtt{DFS}(B.start(),B.end().B.distance())$}
}
\tcc{divide the original bridge $B$ into 2 bridges by $v_m$: $BL$ and $BR$}
$BL$:=$\lbrace B.start(), ...,\overrightarrow{v_m}\rbrace$ \;
$BR$:=$\lbrace \overrightarrow{v_m}, ...,B.end()\rbrace$ \;
\tcc{Return the join of running recursive function on two sub-bridges}
\Return{$\BinaryBridging(BL)\Join\BinaryBridging(BR)$} 
}
\caption{Recursive binary bridging to connect 2 anchor nodes.}
\label{algo:binsearch}
\end{algorithm}

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} 1.1 was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), and \unicycler{} version 0.4.6 %%COMMENT_LC - why just indicate unicyler version and not others? 
using \unicycler{}`s synthetic data set~ \cite{Wick2017unicycler} . 
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{images/FigS2.eps}
 \caption{Alignments of an \emph{Enterobacter~cloacae} reference genome to assembly sequences generated by  (a)~\unicycler{} and (b)~\npgraph{}. The former suggests a structural variant, the latter is virtually an 1-to-1 mapping.}
\end{figure}

\clearpage
\paragraph*{S1 Table.}
\label{S1_Table}
Benchmarking \npgraph{} against \npscarf{} versions, $\mathtt{hybridSPAdes}$ and \unicycler{} hybrid assembler with the synthetic data set.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

%% END MACROS SECTION
\usepackage[normalem]{ulem}
\newcommand\cancel[1]{\sout{#1}}
\newcommand\addnew[1]{\textcolor{red}{#1}}
%% npGraph
\newcommand{\npscarf}{$\mathtt{npScarf}$}
\newcommand{\npscarfg}{$\mathtt{npScarf\_wag}$}
\newcommand{\npreader}{$\mathtt{npReader}$}
\newcommand{\npanalysis}{$\mathtt{npAnalysis}$}
\newcommand{\npbarcode}{$\mathtt{npBarcode}$}
\newcommand{\npgraph}{$\mathtt{npGraph}$}
\newcommand{\canu}{$\mathtt{Canu}$}
\newcommand{\unicycler}{$\mathtt{Unicycler}$}
\newcommand{\spades}{$\mathtt{SPAdes}$}
\newcommand{\hspades}{$\mathtt{hybridSPAdes}$}
\newcommand{\albacore}{$\mathtt{Albacore}$}
\newcommand{\racon}{$\mathtt{Racon}$}
\newcommand{\metrichor}{$\mathtt{Metrichor}$}
\newcommand{\minimap}{$\mathtt{minimap2}$}
\newcommand{\miniasm}{$\mathtt{miniasm}$}
\newcommand{\bwa}{$\mathtt{BWA\text{-}MEM}$}

@@

The first step is to cluster the big anchors (longer than $10$Kbp and in/out degree less than 2) based on their \emph{kmer} coverage.
To achieve this, we applied DBSCAN clustering algorithm \cite{Ester96adensity-based}, which utilises a customised metric function to map contigs into a one-dimensional space.  In order to 
 define a customised metric which is \cancel{sample} \addnew{simple} and fast to calculate, we assumed that a single long contig itself consists of a Poisson distribution of \emph{k-mers} count with the mean approximated by the contig's coverage. 
%The rationale is to approximate a coverage value of an extremely long contig (which can be split into more than 10,000 \emph{k-mers}) with a sampled mean of a Poisson distribution (of \emph{k-mers} count). 
The metric is then determined by a distance function of two Poisson distributions based on Kullback-Leibler divergence (or relative entropy) between the Poisson distribution representing each contig\cite{Kullback1951information}.

@@

Due to false \addnew{positive} alignments from shorter contigs to the long reads, not all of the reported step nodes \cancel{are necessary to be appeared} \addnew{would appear} in the ultimate path resolved by the bridge. 
In most cases, the accumulated score of each step indicates its likelihood to be the true component of the final solution.
For that reason, a strategy similar to binary search\cancel{ing} is employed to find a path across 2 anchors of a bridge as shown in Algorithm~\ref{algo:binsearch}. \addnew{Briefly, it recursively looks for sub-paths connecting pairs of step nodes in-between 2 anchors and return a final continuous path with the highest score.}
\cancel{Before that, }We define Algorithm~\ref{algo:findpath} \addnew{as the \emph{base} sub-rountine} to \cancel{demonstrates the path finding algorithm for} \addnew{find paths between a \emph{connectable} pair of} two \addnew{step} nodes given their estimated distance. In which, function 
\[
\mathtt{shortestTree}(\overrightarrow{vertex},distance) : (V,Z) \rightarrow V^n
\]
from line 3 \cancel{of the algorithm's pseudo code} builds a shortest tree rooted from $\overrightarrow{v}$, following its direction until a distance of approximately $d$ (with a tolerance \cancel{regarding nanopore read} \addnew{for the nanopore} error rate) is reached. This task is implemented based on Dijkstra algorithm.
This tree is used on line 4 and in function $includedIn()$ on line 19 to filter out any node or edge with ending nodes that do not belong to the tree.

@@

\begin{algorithm}[!hpt]
\DontPrintSemicolon
\KwData{Assembly graph $G\{V,E\}$}
\KwIn{Brigde $B:\lbrace\overrightarrow{v_0}, ...\overrightarrow{v_k}, ...,\overrightarrow{v_n}\rbrace$ with $\overrightarrow{v_0}$ and $\overrightarrow{v_n}$ are two anchors, $\{\overrightarrow{v_k}\}, k=1\ldots(n-1)$ are steps in-between}
\KwOut{Set of candidate paths connecting $\overrightarrow{v_0}$ to $\overrightarrow{v_n}$ that maximize the likelihood of the step list.}
\SetKwFunction{BinaryBridging}{BinaryBridging} 
\SetKwProg{Fn}{Function}{:}{}
\Fn{\BinaryBridging{B}}{
\tcc{search for the contig with maximum score from the step list (two ends excluded)}
$m$:=$\argmax_k(\overrightarrow{v_k}.score())$ \;
\tcc{if there is no step in-between, run path finding algorithm above directly and return the result}
\If{$M.size() \equiv 2$}{ 
	\Return{$\mathtt{DFS}(B.start(),B.end().B.distance())$}
}
\tcc{divide the original bridge $B$ into 2 bridges by $v_m$: $BL$ and $BR$}
$BL$:=$\lbrace B.start(), ...,\overrightarrow{v_m}\rbrace$ \;
$BR$:=$\lbrace \overrightarrow{v_m}, ...,B.end()\rbrace$ \;
\tcc{Return the join of running recursive function on two sub-bridges}
\Return{$\BinaryBridging(BL)\Join\BinaryBridging(BR)$} 
}
\caption{Recursive binary bridging to connect 2 anchor nodes.}
\label{algo:binsearch}
\end{algorithm}

@@

% Results and Discussion can be combined.
\section*{Results}
\subsection*{Evaluation using synthetic data}
To evaluate the performance of the method, \npgraph{} 1.1 was tested along with \spades{}, \spades{} hybrid from version 3.13.1, ~\cite{AntipovKM2015}, \npscarf{} (japsa 1.7-02), and \unicycler{} version 0.4.6 using \unicycler{}`s synthetic data set~ \cite{Wick2017unicycler} . 
\addnew{}
The data set is a simulation of Illumina and MinION raw data, generated \emph{in silico} based on available microbial references.
We ran hybrid assembly methods using the entire nanopore data and the reciprocal results were evaluated by QUAST 5.0.2~\cite{Mikheenko2018quast5}.

@@

\begin{figure}[!ht]
 \includegraphics[width=\textwidth]{images/FigS2.eps}
 \caption{Alignments of an \emph{Enterobacter~cloacae} reference genome to assembly sequences generated by  (a)~\cancel{\unicycler{}}\addnew{\npgraph{}} and (b)~\cancel{\npgraph{}}\addnew{\unicycler{}}. The former suggests a structural variant, the latter is virtually an 1-to-1 mapping.}
\end{figure}

\clearpage
\paragraph*{S1 Table.}
\label{S1_Table}
Benchmarking different methods using $\mathtt{\addnew{lrscaf}}$, \npscarf{}, \npgraph{}, $\mathtt{hybridSPAdes}$ and \unicycler{} hybrid assembler with the synthetic data set.

@@

& \addnew{lrscaf} & \addnew{-} & \addnew{-} & \addnew{-} & \addnew{-} & \addnew{-} & \addnew{-} \\*

@@

& \addnew{lrscaf} & \addnew{4113104} & \addnew{6} & \addnew{4002298} & \addnew{0} & \addnew{0.00} & \addnew{0.00}\\*

@@

& \addnew{lrscaf} & \addnew{4119020} & \addnew{12} & \addnew{4002117} & \addnew{0} & \addnew{0.03} & \addnew{0.79}\\* 

@@

& \addnew{lrscaf} & \addnew{3896799} & \addnew{4} & \addnew{3886381} & \addnew{0} & \addnew{0.47} & \addnew{
0.36}\\*

@@

& \addnew{lrscaf} & \addnew{4367657} & \addnew{27} & \addnew{1035967} & \addnew{13} & \addnew{30.11} & \addnew{1.42}\\* 

@@

& \addnew{lrscaf} & \addnew{4732001} & \addnew{16} & \addnew{692794} & \addnew{7} & \addnew{3.42} & \addnew{0.58}\\*

@@

& \addnew{lrscaf} & \addnew{5410854} & \addnew{21} & \addnew{910949} & \addnew{9} & \addnew{13.70} & \addnew{0.87}\\*

@@

& \addnew{lrscaf} & \addnew{5512518} & \addnew{20} & \addnew{777919} & \addnew{26} & \addnew{3.52} & \addnew{0.38}\\*

@@

& \addnew{lrscaf} & \addnew{5823340} & \addnew{38} & \addnew{559414} & \addnew{7} & \addnew{5.92} & \addnew{1.27}\\*

@@

& \addnew{lrscaf} & \addnew{5470282} & \addnew{8} & \addnew{2951488} & \addnew{9} & \addnew{3.04} & \addnew{2.62}\\*

@@

& \addnew{lrscaf} & \addnew{4260869} & \addnew{12} & \addnew{3215170} & \addnew{8} & \addnew{6.69} & \addnew{2.27}\\*

@@

& \addnew{lrscaf} & \addnew{12486884} & \addnew{214} & \addnew{254587} & \addnew{27} & \addnew{16.45} & \addnew{1.97}\\*

@@

& \addnew{lrscaf} & \addnew{4678609} & \addnew{175} & \addnew{376897} & \addnew{26} & \addnew{3.91} & \addnew{2.35}\\*

@@

& \addnew{lrscaf} & \addnew{5266645} & \addnew{128} & \addnew{429353} & \addnew{27} & \addnew{9.72} & \addnew{1.19}\\*

@@

& \addnew{lrscaf} & \addnew{2231488} & \addnew{10} & \addnew{683404} & \addnew{6} & \addnew{11.96} & \addnew{1.37}\\*
